---
title: "COVID-19 Output Report"
author: "Dariusz Adamczyk"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(zoo)
library(knitr)
library(rmarkdown)
library(ggplot2)
library(plotly)
library(caret)
```
## Executive summary
Using a data sample from Tongji Hospital in Wuhan from January 10 to February 18, 2020, the study determined the most relevant parameters to estimate the probability of death for a patient suspected of having COVID-19 virus. The values of the following blood indicators were identified as the parameters most strongly correlated with the probability of death:

- Neutrofils (%) and Neutrofils count 
- Percentage of Lymphocytes
- Albumin
- Prothrombin activity
- High sensitivity C-reactive protein
- D-D Dimer
- Lactate dehydrogenase
- Fibrin degradation products 

The age of the patients is also correlated with patients survivability, however correlation does not imply causation, so therefore the age was decided not to be used in the analysis.

The conducted analysis confirms the key parameter conculsions stated by Li Yan et al. in the article „An interpretable mortality prediction model for COVID-19 patients”. The importance of Lactate dehydrogenase, High sensivity C-reactive protein and Lymphocyte levels was also confirmed in the data presented in the report.

---

Wykorzystując próbkę danych ze szpitala Tongji w Wuhan z okresu od 10 stycznia do 18 lutego 2020 r., w badaniu określono najistotniejsze parametry pozwalające oszacować prawdopodobieństwo zgonu pacjenta podejrzanego o zakażenie wirusem COVID-19. Wartości następujących wskaźników krwi zostały zidentyfikowane jako parametry najsilniej skorelowane z prawdopodobieństwem zgonu:

- Neutrofile
- Odsetek limfocytów
- Albumina
- Aktywność protrombiny
- Białko C-reaktywne o wysokiej czułości
- D-dimery
- Dehydrogenaza mleczanowa
- Produkty degradacji fibryny 

Wiek pacjentów jest również skorelowany z przeżywalnością pacjentów, jednak korelacja nie oznacza związku przyczynowego, dlatego też zdecydowano się nie uwzględniać wieku w analizie.

Przeprowadzona analiza potwierdza kluczowe zbieżności parametrów podane przez Li Yan i wsp. w artykule ""An interpretable mortality prediction model for COVID-19 patients". Znaczenie dehydrogenazy mleczanowej, białka C-reaktywnego o wysokiej czułości oraz poziomu limfocytów zostało również potwierdzone w danych przedstawionych w raporcie.

## Used R libraries

```{r}
sessionInfo()
```

## Code ensuring repeatability of results each time the report runs on the same data

```{r}
rm(list=ls())
```

## Code allowing to load data from the input file

```{r}
setwd("C:/Users/adamc/OneDrive/Desktop/Studia Podyplomowe/Projekt R")

coviddata <- read_excel("wuhan_blood_sample_data_Jan_Feb_2020.xlsx")
```

## Data cleansing code

```{r}
colnames(coviddata)[1] <- "Patient ID"
colnames(coviddata)[2] <- "Date of entry"
colnames(coviddata)[3] <- "Age"
colnames(coviddata)[4] <- "Gender"
colnames(coviddata)[7] <- "Outcome"
colnames(coviddata)[9] <- "Hemoglobin"
colnames(coviddata)[12] <- "Procalcitonin"
colnames(coviddata)[13] <- "Eosinophils"
colnames(coviddata)[16] <- "Albumin"
colnames(coviddata)[17] <- "Basophil"
colnames(coviddata)[21] <- "Monocytes"
colnames(coviddata)[22] <- "Antithrombin"
colnames(coviddata)[24] <- "Indirect bilirubin"
colnames(coviddata)[26] <- "Neutrophils"
colnames(coviddata)[27] <- "Total protein"
colnames(coviddata)[31] <- "Mean corpuscular volume"
colnames(coviddata)[32] <- "Hematocrit"
colnames(coviddata)[33] <- "White blood cell count"
colnames(coviddata)[34] <- "Tumor necrosis factor alpha"
colnames(coviddata)[35] <- "Mean corpuscular hemoglobin concentration"
colnames(coviddata)[36] <- "Fibrinogen"
colnames(coviddata)[39] <- "Lymphocyte count"
colnames(coviddata)[45] <- "Glucose"
colnames(coviddata)[46] <- "Neutrophils count"
colnames(coviddata)[49] <- "Ferritin"
colnames(coviddata)[52] <- "Lymphocyte"
colnames(coviddata)[56] <- "Aspartate aminotransferase"
colnames(coviddata)[59] <- "Calcium"
colnames(coviddata)[62] <- "Platelet large cell ratio"
colnames(coviddata)[65] <- "Monocytes  count"
colnames(coviddata)[67] <- "Globuline"
colnames(coviddata)[68] <- "Gamma-glutamyl transpeptidase"
colnames(coviddata)[70] <- "Basophil count"
colnames(coviddata)[72] <- "Mean corpuscular hemoglobin"
colnames(coviddata)[76] <- "Serum sodium"
colnames(coviddata)[77] <- "Thrombocytocrit"
colnames(coviddata)[79] <- "Glutamic-pyruvid transaminase"
colnames(coviddata)[81] <- "Creatinine"
```

## Section summarizing dataset size and basic statistics

```{r}
print(paste("Number of gathered inputs:",nrow(coviddata),sep=" "))
print(paste("Number of analyzed patients:",max(coviddata$`Patient ID`),sep=" "))
print(paste("Lowest patients' age:",min(coviddata$Age),sep=" "))
print(paste("Mean patients' age:",mean(coviddata$Age),sep=" "))
print(paste("Highest patients' age:",max(coviddata$Age),sep=" "))
```

## Analysis of attribute values

```{r echo=FALSE}
print("Patients age")
summary(coviddata$Age)
hist(coviddata$Age, main="Patients age histogram", xlab="Patients age")

summary(coviddata[-1:-7])
```

## The correlation check section

```{r warning=FALSE}

pearsoncor <- matrix(data=NA, nrow=0, ncol=2)


  for(i in 8:81){
    outcome <- dplyr::pull(coviddata,7)
    analyzed_data <- dplyr::pull(coviddata,i)
    currentfactor <- colnames(coviddata)[i]
    corvalue <- cor.test(outcome, analyzed_data)$estimate
    result <- c(currentfactor, abs(corvalue))
    pearsoncor <- rbind(pearsoncor, result)
    
  }
pearsoncor <- pearsoncor[order(pearsoncor[,2], decreasing=TRUE),]

plot(x <- pearsoncor[1:12,2], main="Most important factors predicting outcome", ylab="Pearson Correlation")

text(pearsoncor[1:12,2], labels=pearsoncor[1:12,1], cex=0.7)

p <- ggplot(coviddata, aes(x=coviddata$Neutrophils, color=factor(coviddata$Outcome))) + geom_histogram(binwidth = 1, fill="beige") + xlab("Neutrophils") + ylab("Outcome") + ggtitle("Impact of Neutrophils on Outcome") + scale_color_manual(labels = c("Survived", "Died"), values = c("darkgreen", "red")) + labs(color="Outcome")
plot(p)

p2 <- ggplot(coviddata, aes(x=coviddata$Lymphocyte, color=factor(coviddata$Outcome))) + geom_histogram(binwidth = 1, fill="beige") + xlab("Lymphocyte") + ylab("Outcome") + ggtitle("Impact of Lymphocyte on Outcome") + scale_color_manual(labels = c("Survived", "Died"), values = c("darkgreen", "red")) + labs(color="Outcome")
plot(p2)

p3 <- ggplot(coviddata, aes(x=coviddata$Albumin, color=factor(coviddata$Outcome))) + geom_histogram(binwidth = 1, fill="beige") + xlab("Albumin") + ylab("Outcome") + ggtitle("Impact of Albumin on Outcome") + scale_color_manual(labels = c("Survived", "Died"), values = c("darkgreen", "red")) + labs(color="Outcome")
plot(p3)

p4 <- ggplot(coviddata, aes(x=coviddata$`Prothrombin activity`, color=factor(coviddata$Outcome))) + geom_histogram(binwidth = 1, fill="beige") + xlab("Prothrombin Activity") + ylab("Outcome") + ggtitle("Impact of Prothrombin Activity on Outcome") + scale_color_manual(labels = c("Survived", "Died"), values = c("darkgreen", "red")) + labs(color="Outcome")
plot(p4)

p5 <- ggplot(coviddata, aes(x=coviddata$`High sensitivity C-reactive protein`, color=factor(coviddata$Outcome))) + geom_histogram(binwidth = 1, fill="beige") + xlab("High sensivity C-reactive protein") + ylab("Outcome") + ggtitle("Impact of High sensivity C-reactive protein on Outcome") + scale_color_manual(labels = c("Survived", "Died"), values = c("darkgreen", "red")) + labs(color="Outcome")
plot(p5)

p6 <- ggplot(coviddata, aes(x=coviddata$`D-D dimer`, color=factor(coviddata$Outcome))) + geom_histogram(binwidth = 1, fill="beige") + xlab("D-D dimer") + ylab("Outcome") + ggtitle("Impact of D-D dimer on Outcome") + scale_color_manual(labels = c("Survived", "Died"), values = c("darkgreen", "red")) + labs(color="Outcome")
plot(p6)

p7 <- ggplot(coviddata, aes(x=coviddata$`Lactate dehydrogenase`, color=factor(coviddata$Outcome))) + geom_histogram(binwidth = 1, fill="beige") + xlab("Lactate dehydrogenase") + ylab("Outcome") + ggtitle("Impact of Lactate dehydrogenase on Outcome") + scale_color_manual(labels = c("Survived", "Died"), values = c("darkgreen", "red")) + labs(color="Outcome")
plot(p7)

p8 <- ggplot(coviddata, aes(x=coviddata$`Neutrophils count`, color=factor(coviddata$Outcome))) + geom_histogram(binwidth = 1, fill="beige") + xlab("Neutrophils count") + ylab("Outcome") + ggtitle("Impact of Neutrophils count on Outcome") + scale_color_manual(labels = c("Survived", "Died"), values = c("darkgreen", "red")) + labs(color="Outcome")
plot(p8)

p9 <- ggplot(coviddata, aes(x=coviddata$`Fibrin degradation products`, color=factor(coviddata$Outcome))) + geom_histogram(binwidth = 1, fill="beige") + xlab("Fibrin degradation products") + ylab("Outcome") + ggtitle("Impact of Fibrin degradation products on Outcome") + scale_color_manual(labels = c("Survived", "Died"), values = c("darkgreen", "red")) + labs(color="Outcome")
plot(p9)
```

## Interactive graph showing the change of selected attributes over time

```{r}

coviddata$Gender[coviddata$Gender==1] <- "Male"
coviddata$Gender[coviddata$Gender==2] <- "Female"

#coviddata$Outcome[coviddata$Outcome==0] <- "Survived"
#coviddata$Outcome[coviddata$Outcome==1] <- "Died"

p10 <- ggplot(coviddata, aes(x=Age, y=Gender, color=Outcome)) + geom_point() + scale_color_distiller() + theme_classic() + theme(legend.title = element_blank())
ggplotly(p10)
```

## Patient survival classifier

```{r}

coviddata.training.indicies <- createDataPartition(coviddata$Outcome, p = 0.80, list = FALSE)
coviddata.training <- coviddata[coviddata.training.indicies,]
coviddata.validation <- coviddata[-coviddata.training.indicies,]

control <- trainControl(method="cv", number=10)
metric <- "Accuracy"

library(MASS)
#set.seed(7)
#fit.lda <- train(Outcome~., data=coviddata.training, method="lda", metric=metric, trControl=control )

#set.seed(7)
#fit.cart <- train(Outcome~., data=coviddata.training, method="cart", metric=metric, trControl=control )

#set.seed(7)
#fit.knn <- train(Outcome~., data=coviddata.training, method="knn", metric=metric, trControl=control )

#set.seed(7)
#fit.svm <- train(Outcome~., data=coviddata.training, method="svm", metric=metric, trControl=control )

#set.seed(7)
#fit.rf <- train(Outcome~., data=coviddata.training, method="rf", metric=metric, trControl=control )
```

## Importance analysis of the attributes of the best model found


The correlation analysis conducted in the report indicates several factors having major influence over predicted patients outcome. The following parameters seems to have the most significance:

- Neutrofils (%) and Neutrofils count 
- Percentage of Lymphocytes
- Albumin
- Prothrombin activity
- High sensitivity C-reactive protein
- D-D Dimer
- Lactate dehydrogenase
- Fibrin degradation products 

The results of data analysis are consistent with the article „An interpretable mortality prediction model for COVID-19 patients”. The proposed estimated outcome algorithm is dependent on three factors:

- Lactate dehydrogenase
- High sensitivity C-reactive protein
- Lymphocyte

All three parameters are among the most significant factors obtained during analysis.


